# (PART\*) Unemployment {-} 

# Unemployment Insurance {#ui}

_Sachin S. Pandya_, 2020-10-03

This Chapter covers how changes in the law of unemployment insurance (UI) affected workers who became unemployed because of COVID-19. As COVID-19 hit the US and spread, and more workers became unemployed as a result, unemployed UI claimants faced the problem of getting and keeping UI benefits. 
Congress also pushed States to relax some UI rules, and to provide aid of its own. In Title II(A) of the [CARES Act, Pub. L. No. 116-136, 134 Stat. 281 (March 27, 2020)](https://www.govinfo.gov/content/pkg/PLAW-116publ136/uslm/PLAW-116publ136.xml), Congress provided for additional UI payments, including a newly-created Federal Pandemic Unemployment Compensation ("FPUC") benefit and a new Pandemic Unemployment Assistance ("PUA") benefit, along with eligibility rules that applied on top of, or as exceptions to, State UI eligibility rules. The result: A complex web of eligibility rules between workers and UI monetary benefits.

## UI Eligibility {#ui_elig}

In the US, there are 53 sets of UI eligibility rules-- one for each State, plus the District of Columbia, Puerto Rico, and the Virgin Islands. At the same time, State UI systems are similar in many ways, in part because to get certain tax advantages, the US Department of Labor (DOL) must certify that each State's UI system meets some basic UI conditions set by federal law. See [42 U.S.C. $\S$ 503](https://www.govinfo.gov/content/pkg/USCODE-2018-title42/html/USCODE-2018-title42-chap7-subchapIII-sec503.htm); [20 C.F.R. $\S$ 604.1 - 604.6](https://www.ecfr.gov/cgi-bin/retrieveECFR?gp=&SID=f6d24b9ec09459c3e1487411a97e871f&mc=true&r=PART&n=pt20.3.604).

Typically, an applicant for UI must leap over several eligibility hurdles to receive benefits. You typically need to have earned a minimum amount from work during a year-long period prior to when you apply for UI. And you must have left your prior job involuntarily ("involuntary separation"). That means that States can deny benefits for (1) voluntarily quitting without a good enough reason ("good cause"); (2) discharge due to job-related misconduct, (3) unemployment because of a labor dispute, and (4) fraud to get or increase benefits [@DOL2019, chap. 5].

If you are eligible, the UI weekly benefit amount is typically some function of your past wages in a "base period" of time -- typically the first four of the last five completed calendar quarters -- as adjusted by minimum and maximum benefit amounts, which vary a lot by State [Figure \@ref(fig:wklybnft); @DOL_sigUI2020, cols. 2-4].


```{r wklybnft, echo=FALSE, message=FALSE, fig.cap = "Min/Max UI Weekly Benefits by State, January 2020"}

library(dplyr)

# https://oui.doleta.gov/unemploy/content/sigpros/2020-2029/January2020.pdf

# State UI Benefit MinMax 2020

load_path <- here::here("data", "state_UI_bnfts_minmax_2020-01.csv")
d <- readr::read_csv(load_path)

d2 <- d %>%
  tidyr::pivot_longer(-state, names_to = "benefit", values_to = "amount") %>%
  arrange(desc(amount))

library(ggplot2)

dotplot <- ggplot(d2, aes(x = amount, y = reorder(state, amount))) + geom_line(aes(group = state)) +
  geom_point(aes(color = benefit)) + 
  theme_minimal() +
  scale_x_continuous(labels = scales::dollar) +
  theme(axis.title = element_blank(),
        axis.text.y = element_text(size=6),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.title = element_blank(),
        legend.justification = c(0, 1), 
        legend.position = c(.1, 1.075),
        legend.background = element_blank(),
        legend.direction="horizontal") + 
  labs(
      caption = "Source: Department of Labor (2020a)"
      )

dotplot


```

Then, once you start getting UI benefits, you've got to do some things to keep on getting them. For the Department of Labor (DOL) to certify a State's UI program, that State's UI law must at least require that, "for regular compensation for any week, a claimant must be able to work, available to work, and actively seeking work." 42 U.S.C. $\S$ 503(a)(12). 

What counts as "able", "available", and "actively seeking" can vary. For example, a State may take an individual as "available" to work if, say, that individual is "on temporary lay-off and is available to work only for the employer that has temporarily laid-off the individual." [20 CFR § 604.5(a)(3)](https://www.ecfr.gov/cgi-bin/retrieveECFR?gp=&SID=f6d24b9ec09459c3e1487411a97e871f&mc=true&r=PART&n=pt20.3.604#se20.3.604_15). Moreover, while an individual must be "available for work," State UI law may limit that availability to only "suitable" work so long as that doesn’t let an individual “withdraw[ ] from the labor market” while receiving UI benefits [20 C.F.R. 604.5(a)(2)]( https://www.ecfr.gov/cgi-bin/text-idx?SID=2564f6f58a76c73ef8fe2c46e4ce4768&mc=true&node=se20.3.604_15&rgn=div8). Most States' UI law limits the availability requirement to "suitable" work, though what counts as "suitable" and as a good-enough reason ("good cause") to refuse suitable work, varies by State [@DOL2019, 5-33 - 5-38]. 

In any case, no State UI system may deny UI benefits to an otherwise eligible individual for refusing to accept "new work . . . if the wages, hours, or _other conditions_ of the work offered are substantially less favorable to the individual than those prevailing for similar work in the locality."  [26 U.S.C. § 3304(a)(5)(B)]( https://www.govinfo.gov/content/pkg/USCODE-2018-title26/html/USCODE-2018-title26-subtitleC-chap23-sec3304.htm) (emphasis added).  The DOL has read "other conditions" in this statute to include "health and safety rules" and "physical conditions such as heat, light and ventilation" [@UIPL4198]. This supports reading "other conditions" to cover conditions relevant to COVID-19 exposure risk at work. But it is likely harder for UI recipients to prove that the offered new work "substantially" differs from the "prevailing" conditions for similar work, in part because some businesses have refused to report COVID-19 cases among their workers to co-employees or the public [e.g. @Dungca2020; @Eidelson2020], let alone whether their workplace injury rates are above the industry average [e.g. @Evans2020].

## The State UI Response

As part of the early response, State governors or agencies began relaxing or suspending some UI eligibility rules for COVID-19-related UI claims -- about half the States had done this by end of March 2020 [@Xie2020, appendix] -- some apparently following the lead of an early DOL guidance [@Pallasch2020]. For example, Ohio's governor declared that, under Ohio UI law, a UI applicant is unemployed and not required to "actively seek" work if the applicant is "requested by a medical professional, local health authority, or employer to be isolated or quarantined as a consequence of COVID-19 even if not actually diagnosed with COV-19," except where the worker has "access to leave benefits from their employer(s)." [Ohio E.O. 2020-03D](https://governor.ohio.gov/wps/portal/gov/governor/media/executive-orders/executive-order-2020-03-d). 

When Congress passed the Families First Coronavirus Response Act, [Pub. L. No. 116-127, 134 Stat. 178 (2020)](https://www.govinfo.gov/content/pkg/PLAW-116publ127/uslm/PLAW-116publ127.xml) in mid-March 2020, Congress told DOL that, for purposes of applying its regular UI certification requirements, the DOL "shall" disregard any State modifications of UI "law and policies with respect to work search, waiting week, good cause, or employer experience rating on an emergency temporary basis as needed to respond to the spread of COVID-19." Act § 4102(b). Thereafter, DOL issued a guidance on what temporary measures States could take. For example, DOL suggested that States modify their UI requirement that any voluntary quits must be for "good cause" to allow workers to leave work "due to reasonable risk of exposure or infection (i.e., self-quarantine) or to care for a family member affected by the virus" [@Pallasch2020, 8].

Thereafter, some States have adopted additional COVID-19 modifications to their UI law. For example, in August 2020, the Nevada legislature directed its UI agency to "establish justifications related to COVID-19 that may constitute good cause for a person to refuse suitable work," and provided a non-exhaustive list of such justifications. Those justifications include an unreasonable risk of workplace COVID-19 exposure of a person who, according to the Centers for Disease Control, falls within a high COVID-19 risk category; the  person  is  caring  for  a  child  who  is  unable  to  attend school or a child care facility because of COVID-19; or the person is at or above age 65. [Nevada, SB3 § 15 (Aug. 6, 2020)](https://www.leg.state.nv.us/App/NELIS/REL/32nd2020Special/Bill/7155/Text).

## FPUC: Federal Pandemic Unemployment Compensation {#fpuc}

Thanks to Title II(A) of the [CARES Act, Pub. L. No. 116-136, § 2104, 134 Stat. 281, 319](https://www.govinfo.gov/content/pkg/PLAW-116publ136/uslm/PLAW-116publ136.xml) (enacted March 2020), certain workers became eligible for Federal Pandemic Unemployment Compensation (FPUC) --  an additional $600 per week for weeks of unemployment ending on or before July 31, 2020, on top of what they could get in State UI benefits. 

Congress initially settled on the flat $600 FPUC amount in the CARES Act, rather than making the FPUC amount depend on a claimant's total weekly UI payment under State law. A plausible reason for a _flat_ amount: it is harder for States to rewrite the software of their antiquated UI claims processing computer systems to implement a FPUC that varied by a claimant's prior wages during their "base period". Adding a flat amount is a lot easier. About a decade earlier, in 2009, 90% of State UI agencies had reported that their benefits or tax systems ran on old hardware and software programming languages, such as COBOL (average age of State UI benefits I.T. system = 22 years) [@NASWA2010, 2]. A decade later, most UI claims in 2019 were filed via the Internet, though in some States (e.g, Kansas, Vermont), many still filed by phone [@DOLfiling]. 

When the CARES Act passed, State UI agencies could not immediately begin processing and paying out FPUC benefits. Some States, such as New Jersey and Connecticut, put out the call for COBOL programmers to help them rewrite their code, while others argued that aging hardware or years of understaffing, not COBOL, was really to blame [@Botella2020; @Hicks2020]. Rhode Island, in contrast, bypassed its legacy system and, within ten days after the CARES Act, built and deployed a separate cloud-based system for collecting, validating, and paying FPUC claims [@Angell2020].

```{r FPUC, eval=FALSE, echo=FALSE}

# ETA 2112 

# c51 Net UI Benefits
# c125 the net amount of FPUC paid (line 42b, column F)

# if no activity, no report filed

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)

data <- read_csv("https://oui.doleta.gov/unemploy/csv/ar2112.csv")
write_csv(data,"ar2112.csv")

library(lubridate)

d <- data %>%
  mutate(report_date = mdy(rptdate)) %>%
  mutate(report_year = year(report_date)) %>%
  rename(stateab = st) %>%
  select(stateab,report_date,report_year,c103, c121, c125) %>%
    filter(stateab != "VI",
         #c10 != 0,
         c125 !=0,
         report_year > 2019
         ) %>%
    mutate(fpuc = c125/1000000000)

fpuc_plot <- ggplot(d,aes(x = fpuc, y = reorder(stateab,fpuc), 
                     fill=as.factor(report_date),color=as.factor(report_date))) +
              geom_point() +
              theme_minimal() + 
              ylab("") + 
              xlab("Billions of dollars") + 
              guides(color = FALSE, fill = FALSE) + 
			  labs(
				title = "Net Federal Pandemic Unemployment Compensation Paid",
				caption = "Source: ETA 2112"
				)

fpuc_plot

#save_path <- here::here("figures_final")
#ggsave("fpuc_plot.png", path = save_path)

```

By July 2020, the flat \$600 FPUC amount was still politically controversial. Some politicians complained that the $600 amount was too large, because in some cases, the total weekly UI benefit exceeded wage replacement, i.e., what what they would have received in weekly wages had they gone back to work. Research at the time, however, found little evidence that high UI amounts affected return to work [@Bartik2020; @Altonji2020]. As politicians debated, DOL asked state workforce agencies (which administer State UI programs), in a July 13 survey from DOL's UI Administrator, how long it would take to implement a revised FPUC tied to a claimant's past wages, adding “I know from our conversations with a number of you that these are heavy lifts, and we want to be able to quantify that appropriately.”[quoted in @Penn2020]

The \$600 FPUC provision of the CARES Act only applied to "weeks of unemployment . . . ending on or before July 31, 2020." CARES Act § 2104(e).  Accordingly, for individuals getting UI payments for weeks of unemployment past July 31, those individuals do not get any FPUC amount.

## PUA: Pandemic Unemployment Assistance {#pua}

The CARES Act also separately provided a Pandemic Unemployment Assistance benefit (PUA) for someone who wasn't eligible for regular or extended UI benefits under State or Federal law, including someone who had "exhausted" those benefits. [CARES Act § 2102(a)(3)(A)(i)](https://www.govinfo.gov/content/pkg/PLAW-116publ136/uslm/PLAW-116publ136.xml).  If you qualified for PUA, you would get a weekly amount equal to the State UI amount _plus_ the flat FPUC amount that you would have been otherwise entitled to receive. CARES Act § 2102(d).  Because the FPUC only covers weeks of unemployment up through July 31, 2020, the PUA amounts for weeks after July 31 are \$600 less.  

To qualify for PUA, workers must self-certify that they were "otherwise able to work and available for work" under their State's UI law, but they were "unemployed, partially unemployed, or unable or unavailable to work because" of one of several qualifying conditions, CARES Act § 2102(a)(3)(A)(ii)(I). While an individual is eligible for PUA even if self-employed, seeking only part-time employment, lacked sufficient work history, "or otherwise would not qualify for" regular or extended UI benefits under State or Federal law, CARES Act § 2102(a)(3)(A)(ii)(II), an individual is ineligible for PUA if they could "telework with pay" or was receiving "paid sick leave or other paid leave benefits," CARES Act § 2102(3)(B).

PUA qualifying conditions included that the individual is caring for a family or household member diagnosed with COVID-19. CARES Act § 2102(a)(3)(A)(ii)(I)(aa)-(cc). Another PUA qualifying condition is that 

>a child or other person in the household for which the individual has primary caregiving responsibility is unable to attend school or another facility that is closed as a direct result of the COVID–19 public health emergency and such school or facility care is required for the individual to work;

CARES Act § 2102(a)(3)(A)(ii)(I)(dd). For this qualifying condition, DOL treats a school as "closed" even if the school system is only providing online instruction _or_ a combination of online and in-person instruction, because the students cannot be physically present at the school on the days of online instruction. But if the school lets the individual seeking PUA choose between full-time in-person instruction and remote learning for the student, the school doesn't count as "closed" [@UIPL1620].

Other PUA qualifying conditions include that the individual had to "quit his or her job as a direct result of COVID-19"; "the individual's place of employment is closed as a direct result of the COVID-19 public health emergency"; or the individual "meets any additional criteria established by the Secretary [of Labor] for unemployment assistance under this section." CARES Act § 2102(a)(3)(A)(ii)(I)(ii)-(kk). 

In early April 2020, DOL identified "additional criteria" to address PUA eligibility for independent contractors:

>[A]n individual who works as an independent contractor with reportable income may also qualify for PUA benefits if he or she is unemployed, partially employed, or unable or unavailable to work because the COVID-19 public health emergency has severely limited his or her ability to continue performing his or her customary work activities, and has thereby forced the individual to suspend such activities.

For example, even if a driver for a "ridesharing service" is treated as an independent contractor, and thus has no "place of employment" that closed due to COVID-19, the driver may still qualify for PUA "if he or she has been forced to suspend operations as a direct result of the COVID19 public health emergency, such as if an emergency state or municipal order restricting movement makes continued operations unsustainable" [@UIPL1620a, I-6].

Since March, millions of people have received PUA (Figure \@ref(fig:pua) ).


```{r pua, echo=FALSE, message = FALSE, warning = FALSE, fig.cap = "PUA Eligible Claimants by Month"}

# ETA 902P

#data <- readr::read_csv("https://oui.doleta.gov/unemploy/csv/ap902.csv")
save_path <- here::here("data","ap902.csv")
#readr::write_csv(data,save_path)

library(tidyverse)

data <- readr::read_csv(save_path)

library(lubridate)

d <- data %>%
  mutate(report_date = mdy(rptdate)) %>%
  rename(stateab = st,
         app_total = c1,
         elig_total = c2,
         app_se = c7,
         elig_se = c8,
         f_pymnt = c3,
         wks_clm = c4,
         wks_cmp = c5,
         amt_cmp = c6)

d <- d %>%
  select (stateab, report_date,
          app_total, elig_total,
          app_se, elig_se,
          f_pymnt, wks_clm,
          wks_cmp, amt_cmp
          ) %>%
  mutate(month_date = floor_date(report_date, "month"))

d_all <- d %>%
  filter (month_date != "2020-09-01",
          month_date !="2020-02-01",
          month_date != "2020-01-01") %>%
  group_by(month_date) %>%
  summarise(across(where(is.numeric), sum))

d_elig <- d_all %>%
  select(month_date,elig_total,elig_se) %>%
  pivot_longer(cols = starts_with("elig"), names_prefix = "elig_", names_to = "type", values_to = "count")

plot_elig <- ggplot(data = d_elig, aes(x = month_date, 
                                 y = count/1000000, 
                                 color = type,
                                 shape = type)) + 
        geom_line() + geom_point() +
        theme_minimal() + ylab("Count (millions)") + xlab("") +  
            theme(axis.text.x = element_text(angle = 90, vjust=0.5), 
                  legend.position = "bottom") +
            labs(title = "PUA Eligible Applicants by Month",
                 subtitle = "March - August 2020",
                  caption = "Data Source: ETA 902P") +
             scale_color_discrete(name = "Claimant Type", 
                                 labels=c("self-employed", "all")) +
             scale_shape_discrete(name = "Claimant Type", 
                                 labels=c("self-employed", "all"))
            
plot_elig

```

Not surprisingly, States vary as to the claimants they've determined eligible for PUA, and how many of those are self-employed (Figure \@ref(fig:puastate)).

```{r puastate, echo = FALSE, message = FALSE, fig.cap = "PUA Recipients, Selected States"}

d_state <- d %>%
  select(stateab,month_date,elig_total,elig_se) %>%
  pivot_longer(cols = starts_with("elig"), names_prefix = "elig_", names_to = "type", values_to = "count")

d_state <- d_state %>%
  filter(stateab == "CA" |
         stateab == "PA" |
         stateab == "OH" |   
         stateab == "NY" |
         stateab == "GA" |
         stateab == "IL" |
         stateab == "NC" |
         stateab == "NJ" |
         stateab == "TX")

plot_state <- ggplot(data = d_state, aes(x = month_date, 
                                 y = count/1000, 
                                 color = type,
                                 shape = type)) + 
        geom_line() + geom_point() +
        theme_minimal() + ylab("Count (1000s)") + xlab("") +  
            theme(axis.text.x = element_text(angle = 90, vjust=0.5), 
                  legend.position = "bottom") +
            labs(title = "PUA Eligible Applicants",
                 subtitle = "Selected States, March - August 2020",
                  caption = "Data Source: ETA 902P") +
             scale_color_discrete(name = "Claimant Type", 
                                 labels=c("self-employed", "all")) +
             scale_shape_discrete(name = "Claimant Type", 
                                 labels=c("self-employed", "all")) +
              facet_wrap(~ stateab)            
      
plot_state

```

PUA benefits only cover "weeks of unemployment, partial unemployment, or inability to work caused by COVID–19" up through December 31, 2020. CARES Act § 2102(c)(1)(A).
